Using occupancy theory and Bayesian models to estimate otter populations in southeastern Minnesota

### 2010-03-04 Presentation.tex
Massive anal reorganization, added psi estimate graph.

\documentclass{beamer}
\usetheme{}
\title{Using occupancy theory and Bayesian models to estimate otter populations
in southeastern Minnesota}
\author[Aing, Halls, Oken]{Chrisna Aing, Sarah Halls, Kiva Oken}
\institute{Carleton College}
\usepackage{graphics,graphpap}
\usepackage{color}
\usepackage{multirow}
%\includeonlyframes{currentslide}

\begin{document}
{
	\usebackgroundtemplate{
	\includegraphics[width=\paperwidth]{ottertracks1.jpg}}
	\begin{frame}
		\titlepage
	\end{frame}
}

@@

\section{The data and questions}
\begin{frame}{Otters}
	\begin{columns}
		\column{5cm}
		\begin{itemize}[<+->]
			\item Otters endemic to all of Minnesota
			\item Indicator species for quality of aquatic habitat
			\item Populations crashed, now rebounding
			\item In SE Minnesota interest in resuming trapping
			\item Government needs tools to monitor changes in populations
		\end{itemize}
		\column{5cm}
		\resizebox{5cm}{!}{\includegraphics{cuteOtter1.pdf}}
	\end{columns}
\end{frame}

\begin{frame}{Map of area}
	\begin{picture}(310,190)(9,7.5)
		\put(20,-5){\resizebox{10cm}{!}{\includegraphics{map.pdf}}}
		%\graphpaper(0,0)(310,190)
        		\only<2>{\put(164,48){\circle{30}}}
   	\end{picture}
\end{frame}

\begin{frame}{Collecting data}
	\begin{columns}
		\column{5cm}
		\resizebox{5cm}{!}{\includegraphics{cuteOtter2.jpg}}
		\column{5cm}
		\begin{itemize}[<+->]
			\item Researchers flew over rivers during winters of 2003, 2004
			\item Flew one, two, or three days after snow events
			\item Marked GPS coordinates of tracks
			\item Rivers divided into 402m, 804m, or 1608m plots
			\item Presence of track in each plot determined
			\item Also examined data set with every other plot excluded
			(Alternating Plots)
		\end{itemize}
	\end{columns}
\end{frame}

{
	\usebackgroundtemplate{
	\includegraphics[width=\paperwidth]{SnowTracks12.jpg}}
	\begin{frame}{}
	\end{frame}
}

{
	\usebackgroundtemplate{
	\includegraphics[width=\paperwidth]{SNOWTRACKS9.jpg}}
	\begin{frame}
	\end{frame}
}

{
\usebackgroundtemplate{\includegraphics[width=\paperwidth]{data.pdf}}
\begin{frame}
    \begin{picture}(310,190)(9,7.5)
        \linethickness{1.5pt}
        \only<2->{\color<2>{red}\put(50,120){\framebox(290,15){}}}
        \only<3->{\color<3>{red}\put(150,-20){\framebox(15,80){}}}
        \only<4->{\color<4>{red}\put(150,-35){\framebox(103,30){}}}
    \end{picture}
\end{frame}
}

\section{Occupancy Theory}
\begin{frame}{Definitions of occupancy}
	\begin{itemize}
		\item The probability that a randomly selected site or sampling unit in
		an area of interest is occupied by a species
		\item The proportion of area, patches, or sampling units that is
		occupied
	\end{itemize}
	(MacKenzie, et al. 2006)
	\begin{center}
		\resizebox{5cm}{!}{\includegraphics{cuteOtter3.jpg}}
	\end{center}
\end{frame}

@@

\begin{frame}{Assumptions of occupancy models}
	\begin{columns}
		\column{5cm}
		\begin{itemize}
			\item No false positives
			\item Presence and detection probabilities are constant across sites
			and surveys (or can be modeled by covariates)
			\item Independence of detection between sites
			\item Survey is conducted on a closed population
		\end{itemize}
		\column{5cm}
		\resizebox{5cm}{!}{\includegraphics{cuteOtter4.jpg}}
	\end{columns}
\end{frame}

\begin{frame}{Types of occupancy models}
	\begin{itemize}
		\item Single species, single season
		\item Hierarchical models
	\end{itemize}
	\begin{center}
		\resizebox{7cm}{!}{\includegraphics{../SimpleHierarchicalModel.pdf}}
	\end{center}
\end{frame}

\begin{frame}{Parameters}
	\begin{itemize}
		\item $\psi=$ The probability a site is occupied
		\item $\theta=$ The probability a track is laid down in a site on a
		given day, given that the site is occupied
		\item $p= $ The probability an observer sees a track given that a track
		has been made
	\end{itemize}
	\begin{center}
		\resizebox{7cm}{!}{\includegraphics{../SimpleModelWithParams.pdf}}
	\end{center}
\end{frame}

\section{Exploring the data}
\begin{frame}{Effects of multiple observers}
	\begin{itemize}
		\item In 2003, three different observers
		\item Detection rates varied among the observers
	\end{itemize}
	\begin{center}
		\resizebox{5cm}{!}{\includegraphics{wwObsPlot.pdf}}
	\end{center}
	\begin{itemize}
		\item Solution: use observer indicators as covariates for detection
		probability
	\end{itemize}
\end{frame}

\begin{frame}{Correlation between flights}
	\begin{center}
		\resizebox{7cm}{!}{\includegraphics{dayEventGraphs.pdf}}
	\end{center}
	\begin{itemize}
		\item Also looked at correlation between flights
		\begin{itemize}
			\item Very low, overall
			\item Slightly higher during the same snow event or day
		\end{itemize}
		\item Solution: include false detection in model
	\end{itemize}
\end{frame}

\begin{frame}{Hierarchical model with false detection}
	\begin{center}
		\resizebox{10cm}{!}{\includegraphics{../ModelWithE.pdf}}
	\end{center}
\end{frame}

\begin{frame}{Hierarchical model with false detection}
	\begin{center}
		\resizebox{10cm}{!}{\includegraphics{../ModelWithEAndParams.pdf}}
	\end{center}
\end{frame}

@@

\begin{frame}{Independence, continued}
	\begin{center}
		\includegraphics[scale=0.4]{independenceBoxplots.pdf}
	\end{center}
	\begin{itemize}
		\item Solution: keep issue in mind, include spatially dependent models
	\end{itemize}
\end{frame}

@@

\begin{frame}{Convergence}
	\begin{center}
		\resizebox{8.5cm}{!}{\includegraphics{historyConvergence.pdf}} \\
		\includegraphics[scale=0.8]{distnConvergence.pdf} \\
	\end{center}
\end{frame}

\begin{frame}{Convergence}
	\begin{center}
		\resizebox{8.5cm}{!}{\includegraphics{historyNoConvergence.pdf}} \\
		\includegraphics[scale=0.8]{distnNoConvergence.pdf} \\
	\end{center}
\end{frame}

\begin{frame}{Issue with false detection}
	\begin{columns}
		\column{5cm}
		\begin{itemize}
			\item $(p=x, E=y, \psi=z)$ equivalent in terms of likelihood to
			$(p=y,E=x,\psi=1-z)$.
			\item Leads to bimodal posterior distributions.
			\item Solution: restrict $p$ to interval $(0.6,1)$, $E$ to interval
			$(0,0.4)$.
		\end{itemize}
		\column{5cm}
		\resizebox{5cm}{!}{\includegraphics{cuteOtter6.jpg}}
	\end{columns}
\end{frame}

%This slide for DNR only
%\begin{frame}{Bayesian model specifics}
%	\begin{itemize}
%		\item $\psi,\theta \sim \text{Uniform}(0,1)$
%		\item $p[\text{obs}] \sim \text{Beta}(a,b)$
%		\begin{itemize}
%			\item $a, b \sim \text{Gamma}(1,0.1)$
%			\item E similar
%			\item $p$ and $E$ had restricted distributions
%		\end{itemize}
%		\item Burn-in period 10,000 iterations
%		\item Covariates: observer, days since snow
%	\end{itemize}
%\end{frame}

@@

\begin{frame}{Testing the model, continued}
	\begin{center}
	\includegraphics[scale=0.5]{../SimulatedData.pdf}
	\end{center}
\end{frame}

@@

%DNR only slide
%\begin{frame}{CAR model specifics}
%	\begin{itemize}
%		\item $\tau^2 =$ conditional variance
%		\item $(\tau^2)^{-1} \sim \text{Gamma}(0.5,0.0005)$
%		\item $\text{logit}(\theta_s)=\alpha_0+\alpha_{1_s}$
%		\item $\alpha_0 \sim$ Improper uniform
%		\item $\alpha_1 \sim$ Gaussian CAR
%	\end{itemize}
%\end{frame}

@@

\begin{frame}{}
    % Mississippi River!
	\small{
    \begin{table}
    \begin{center}
    \begin{tabular}{|l|l|l|l|}
        \hline
        \multicolumn{4}{|l|}{\textbf{Mississippi River, 2003}} \\
        \hline
            & 402m & 804m & 1608m \\
        \hline
        \multirow{2}{*}{E[Danny]}
            & 0.01954 & 0.04808 & 0.06885 \\
            & (0.006515, 0.03663) & (0.01874, 0.08187) & (0.01778, 0.1383) \\
        \hline
        \multirow{2}{*}{E[John]}
            & 0.05216 & 0.07771 & 0.13650 \\
            & (0.027710, 0.07745) & (0.03881, 0.12100) & (0.03449, 0.2330) \\
        \hline
        \multirow{2}{*}{E[Tom]}
            & 0.02668 & 0.06140 & 0.10230 \\
            & (0.011990, 0.04622) & (0.02789, 0.10390) & (0.04287, 0.1796) \\
        \hline
        \multirow{2}{*}{p[Danny]}
            & 0.65240 & 0.67340 & 0.74960 \\
            & (0.548300, 0.75990) & (0.56760, 0.78290) & (0.60510, 0.8910) \\
        \hline
        \multirow{2}{*}{p[John]}
            & 0.67770 & 0.77580 & 0.82540 \\
            & (0.583500, 0.77280) & (0.66060, 0.88690) & (0.71390, 0.9326) \\
        \hline
        \multirow{2}{*}{p[Tom]}
            & 0.53670 & 0.54220 & 0.65330 \\
            & (0.501200, 0.61300) & (0.50160, 0.62850) & (0.51530, 0.8104) \\
        \hline
        \multirow{2}{*}{\(\psi\)}
            & 0.59380 & 0.69370 & 0.75820 \\
            & (0.439000, 0.75990) & (0.51800, 0.87380) & (0.54110, 0.9300) \\
        \hline
        \multirow{2}{*}{\(\theta\)}
            & 0.15280 & 0.20730 & 0.27690 \\
            & (0.111900, 0.19780) & (0.14580, 0.27390) & (0.19030, 0.3716) \\
        \hline
    \end{tabular}
    \end{center}
	\end{table} 
	}
\end{frame}

% Add slide with example of estimates for our actual data, one river/year?

\begin{frame}{Summary and interpretation of estimates}
	\begin{itemize}
		\item Estimates for all parameters increase with plot size
		\item Possible hypotheses:
		\begin{itemize}
			\item \(\psi\) and \(\theta\)
			\item \(p\) and \(E\)
		\end{itemize}
		\item Used data from all plots, not alternating plots
		\begin{itemize}
			\item More data more useful than more independence
		\end{itemize}
	\end{itemize}
	\begin{center}
		\includegraphics[scale=0.5]{Aggregation.pdf}
	\end{center}
\end{frame}

@@

\begin{frame}{Acknowledgments}
	\begin{itemize}
		\item Bob Dobrow, adviser
		\item John Fieberg, MN DNR
		\item John Erb, MN DNR
	\end{itemize}
	\begin{center}
		\resizebox{7cm}{!}{\includegraphics{cuteOtter10.jpg}}
	\end{center}
\end{frame}

\begin{frame}{Questions?}
	\begin{center}
	\includegraphics[scale=0.7]{cuteOtter7.jpg}
	\end{center}
\end{frame}

>>>>

\documentclass{beamer}
\usetheme{}
\title{Using occupancy theory and Bayesian models to estimate otter populations
in southeastern Minnesota}
\author[Aing, Halls, Oken]{Chrisna Aing, Sarah Halls, Kiva Oken}
\institute{Carleton College}
\usepackage{graphics,graphpap}
\usepackage{color}
\usepackage{multirow}
\usepackage[T1]{fontenc}
\usepackage{fix-cm}
%\includeonlyframes{currentslide}

\begin{document}
{
	\usebackgroundtemplate{
	\includegraphics[width=\paperwidth]{Figures/Pictures/ottertracks1.jpg}}
	\begin{frame}
		\titlepage
	\end{frame}
}

@@

\section{The data and questions}
\begin{frame}{Otters}
	\begin{columns}
		\column{5cm}
		\begin{itemize}[<+->]
			\item Otters endemic to all of Minnesota
			\item Indicator species for quality of aquatic habitat
			\item Populations crashed, now rebounding
			\item In SE Minnesota interest in resuming trapping
			\item Government needs tools to monitor changes in populations
		\end{itemize}
		\column{5cm}
		\resizebox{5cm}{!}{\includegraphics{Figures/Pictures/cuteOtter1.pdf}}
	\end{columns}
\end{frame}

\begin{frame}{Map of area}
	\begin{picture}(310,190)(9,7.5)
		\put(20,-5){\resizebox{10cm}{!}{\includegraphics{Figures/map.pdf}}}
		\only<2>{\put(164,48){\circle{30}}}
	\end{picture}
\end{frame}

\begin{frame}{Collecting data}
	\begin{columns}
		\column{5cm}
		\resizebox{5cm}{!}{\includegraphics{Figures/Pictures/cuteOtter2.jpg}}
		\column{5cm}
		\begin{itemize}[<+->]
			\item Researchers flew over rivers during winters of 2003, 2004
			\item Flew one, two, or three days after snow events
			\item Marked GPS coordinates of tracks
			\item Rivers divided into 402m, 804m, or 1608m plots
			\item Presence of track in each plot determined
			\item Also examined data set with every other plot excluded
			(Alternating Plots)
		\end{itemize}
	\end{columns}
\end{frame}

{
	\usebackgroundtemplate{
	\includegraphics[width=\paperwidth]{Figures/Pictures/SnowTracks12.jpg}}
	\begin{frame}{}
	\end{frame}
}

{
	\usebackgroundtemplate{
	\includegraphics[width=\paperwidth]{Figures/Pictures/SNOWTRACKS9.jpg}}
	\begin{frame}
	\end{frame}
}

{
\usebackgroundtemplate{\includegraphics[width=\paperwidth]{Figures/data.pdf}}
\begin{frame}
    \begin{picture}(310,190)(9,7.5)
        \linethickness{1.5pt}
        \only<2->{\color<2>{red}\put(50,120){\framebox(290,15){}}}
        \only<3->{\color<3>{red}\put(150,-20){\framebox(15,80){}}}
        \only<4->{\color<4>{red}\put(150,-35){\framebox(103,30){}}}
    \end{picture}
\end{frame}
}

\section{Occupancy Theory}
\begin{frame}{Definitions of occupancy}
	\begin{itemize}
		\item The probability that a randomly selected site or sampling unit in
		an area of interest is occupied by a species
		\item The proportion of area, patches, or sampling units that is
		occupied
	\end{itemize}
	(MacKenzie, et al. 2006)
	\begin{center}
		\resizebox{5cm}{!}{\includegraphics{Figures/Pictures/cuteOtter3.jpg}}
	\end{center}
\end{frame}

@@

\begin{frame}{Assumptions of occupancy models}
	\begin{columns}
		\column{5cm}
		\begin{itemize}
			\item No false positives
			\item Presence and detection probabilities are constant across sites
			and surveys (or can be modeled by covariates)
			\item Independence of detection between sites
			\item Survey is conducted on a closed population
		\end{itemize}
		\column{5cm}
		\resizebox{5cm}{!}{\includegraphics{Figures/Pictures/cuteOtter4.jpg}}
	\end{columns}
\end{frame}

\begin{frame}{Types of occupancy models}
	\begin{itemize}
		\item Single species, single season
		\item Hierarchical models
	\end{itemize}
	\begin{center}
		\resizebox{7cm}{!}{
		\includegraphics{Figures/Diagrams/SimpleHierarchicalModel.pdf}}
	\end{center}
\end{frame}

\begin{frame}{Parameters}
	\begin{itemize}
		\item $\psi=$ The probability a site is occupied
		\item $\theta=$ The probability a track is laid down in a site on a
		given day, given that the site is occupied
		\item $p= $ The probability an observer sees a track given that a track
		has been made
	\end{itemize}
	\begin{center}
		\resizebox{7cm}{!}{
		\includegraphics{Figures/Diagrams/SimpleModelWithParams.pdf}}
	\end{center}
\end{frame}

\section{Exploring the data}
\begin{frame}{Effects of multiple observers}
	\begin{itemize}
		\item In 2003, three different observers
		\item Detection rates varied among the observers
	\end{itemize}
	\begin{center}
		\resizebox{5cm}{!}{\includegraphics{Figures/R/wwObsPlot.pdf}}
	\end{center}
	\begin{itemize}
		\item Solution: use observer indicators as covariates for detection
		probability
	\end{itemize}
\end{frame}

\begin{frame}{Correlation between flights}
	\begin{center}
		\resizebox{7cm}{!}{\includegraphics{Figures/R/dayEventGraphs.pdf}}
	\end{center}
	\begin{itemize}
		\item Also looked at correlation between flights
		\begin{itemize}
			\item Very low, overall
			\item Slightly higher during the same snow event or day
		\end{itemize}
		\item Solution: include false detection in model
	\end{itemize}
\end{frame}

\begin{frame}{Hierarchical model with false detection}
	\begin{center}
		\resizebox{10cm}{!}{\includegraphics{Figures/Diagrams/ModelWithE.pdf}}
	\end{center}
\end{frame}

\begin{frame}{Hierarchical model with false detection}
	\begin{center}
		\resizebox{10cm}{!}{
		\includegraphics{Figures/Diagrams/ModelWithEAndParams.pdf}}
	\end{center}
\end{frame}

@@

\begin{frame}{Independence, continued}
	\begin{center}
		\includegraphics[scale=0.4]{Figures/R/independenceBoxplots.pdf}
	\end{center}
	\begin{itemize}
		\item Solution: keep issue in mind, include spatially dependent models
	\end{itemize}
\end{frame}

@@

\begin{frame}{Convergence}
	\begin{center}
		\resizebox{8.5cm}{!}{
		\includegraphics{Figures/R/historyConvergence.pdf}} \\
		\includegraphics[scale=0.8]{Figures/R/distnConvergence.pdf} \\
	\end{center}
\end{frame}

\begin{frame}{Convergence}
	\begin{center}
		\resizebox{8.5cm}{!}{
		\includegraphics{Figures/R/historyNoConvergence.pdf}} \\
		\includegraphics[scale=0.8]{Figures/R/distnNoConvergence.pdf} \\
	\end{center}
\end{frame}

\begin{frame}{Issue with false detection}
	\begin{columns}
		\column{5cm}
		\begin{itemize}
			\item $(p=x, E=y, \psi=z)$ equivalent in terms of likelihood to
			$(p=y,E=x,\psi=1-z)$.
			\item Leads to bimodal posterior distributions.
			\item Solution: restrict $p$ to interval $(0.6,1)$, $E$ to interval
			$(0,0.4)$.
		\end{itemize}
		\column{5cm}
		\includegraphics[width=5cm]{Figures/Pictures/cuteOtter6.jpg}
	\end{columns}
\end{frame}

% DNR-only slide!
%\begin{frame}{Bayesian model specifics}
%	\begin{itemize}
%		\item $\psi,\theta \sim \text{Uniform}(0,1)$
%		\item $p[\text{obs}] \sim \text{Beta}(a,b)$
%		\begin{itemize}
%			\item $a, b \sim \text{Gamma}(1,0.1)$
%			\item E similar
%			\item $p$ and $E$ had restricted distributions
%		\end{itemize}
%		\item Burn-in period 10,000 iterations
%		\item Covariates: observer, days since snow
%	\end{itemize}
%\end{frame}

@@

\begin{frame}{Testing the model, continued}
	\begin{center}
		\includegraphics[width=8.5cm]{Figures/Diagrams/SimulatedData.pdf}
	\end{center}
\end{frame}

@@

% DNR-only slide!
%\begin{frame}{CAR model specifics}
%	\begin{itemize}
%		\item $\tau^2 =$ conditional variance
%		\item $(\tau^2)^{-1} \sim \text{Gamma}(0.5,0.0005)$
%		\item $\text{logit}(\theta_s)=\alpha_0+\alpha_{1_s}$
%		\item $\alpha_0 \sim$ Improper uniform
%		\item $\alpha_1 \sim$ Gaussian CAR
%	\end{itemize}
%\end{frame}

@@

\begin{frame}{Summary and interpretation of estimates}
	\begin{center}
		\includegraphics[width=10cm]{Figures/Graphs/2003Psi.pdf}
	\end{center}
\end{frame}

\begin{frame}{Summary and interpretation of estimates}
	\begin{itemize}
		\item Estimates for all parameters increase with plot size
		\item Possible hypotheses:
		\begin{itemize}
			\item \(\psi\) and \(\theta\)
			\item \(p\) and \(E\)
		\end{itemize}
		\item Used data from all plots, not alternating plots
		\begin{itemize}
			\item More data more useful than more independence
		\end{itemize}
	\end{itemize}
	\begin{center}
		\includegraphics[width=10cm]{Figures/Diagrams/Aggregation.pdf}
	\end{center}
\end{frame}

@@

\begin{frame}{Acknowledgments}
	\begin{itemize}
		\item Bob Dobrow, adviser
		\item John Fieberg, MN DNR
		\item John Erb, MN DNR
	\end{itemize}
	\begin{center}
		\includegraphics[width=7cm]{Figures/Pictures/cuteOtter10.jpg}
	\end{center}
\end{frame}

\begin{frame}{Questions?}
	\begin{center}
		\includegraphics[width=10cm]{Figures/Pictures/cuteOtter7.jpg}
	\end{center}
\end{frame}
